---
title: "finalproject"
output: html_document
date: "2024-12-03"
---

```{r}
data <- read.csv("kaggle_survey_2020_responses.csv", stringsAsFactors = FALSE)
data <-data[-1,]
data
```


Q1
### Methods
The dataset provides information on respondents' age, gender, country, and highest level of education, alongside their reported income levels. Compensation data was cleaned and transformed by converting ranges into random values within specified bounds to enhance variability. Rows with missing or null compensation values were excluded to maintain data integrity. To explore trends, the data was aggregated and visualized using box plots and bar charts to examine correlations between income and demographic categories such as age, gender, country, and education.

### Analysis


```{r}
selected_data <- data %>%
  select(Q1, Q2, Q3, Q4, Q24) %>%
  rename(
    Age = Q1,
    Gender = Q2,
    Country = Q3,
    Education = Q4,
    Compensation = Q24
  )

# Clean and preprocess the data
# Convert compensation ranges into numeric midpoints
selected_data$Compensation <- gsub("[^0-9\\-]", "", selected_data$Compensation) %>%
  sapply(function(x) {
    if (grepl("-", x)) {
      bounds <- as.numeric(unlist(strsplit(x, "-")))
      return(runif(1, min = bounds[1], max = bounds[2]))
    } else if (x != "") {
      return(as.numeric(x))
    } else {
      return(NA)
    }
  })

selected_data

```
```{r}
# Remove rows with missing values in critical columns
selected_data <- selected_data %>%
  filter(!is.na(Age) & !is.na(Gender) & !is.na(Country) &
           !is.na(Education) & !is.na(Compensation))

# Calculate average income by age, gender, and country
avg_income_by_demo <- selected_data %>%
  group_by(Age, Gender, Country) %>%
  summarise(Average_Compensation = mean(Compensation, na.rm = TRUE), .groups = "drop")

# Calculate average income by education
avg_income_by_edu <- selected_data %>%
  group_by(Education) %>%
  summarise(Average_Compensation = mean(Compensation, na.rm = TRUE), .groups = "drop")

# Visualize: Income Distribution by Education Level
ggplot(selected_data, aes(x = Education, y = Compensation, fill = Gender)) +
  geom_boxplot() +
  labs(title = "Income Distribution by Education Level and Gender",
       x = "Education Level",
       y = "Compensation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Visualize: Average Income by Country
ggplot(avg_income_by_demo, aes(x = reorder(Country, -Average_Compensation), y = Average_Compensation, fill = Gender)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Average Income by Country and Gender",
       x = "Country",
       y = "Average Compensation") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# Visualize: Average Income by Age
ggplot(avg_income_by_demo, aes(x = Age, y = Average_Compensation, fill = Gender)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Average Income by Age and Gender",
    x = "Age",
    y = "Average Compensation",
    fill = "Gender"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The findings reveal distinct income patterns across demographic groups. Age strongly correlates with income, as older age groups (e.g., 55-59, 70+) generally report higher average compensation, likely reflecting accumulated experience. Gender disparities in income are evident, with men earning higher compensation on average across most age and education levels. The role of education is significant, as advanced degrees like doctoral and professional degrees correlate with higher median incomes. However, the variability within each educational category suggests other influencing factors. Country-level differences show significant disparities, with respondents from developed regions (e.g., Switzerland, the United States) reporting higher incomes compared to those from developing regions.

### Conclusion
The analysis demonstrates the substantial impact of age, gender, country, and education on income levels among data science professionals. Advanced education and experience play a critical role in achieving higher compensation, though gender disparities persist. Regional economic conditions further amplify these differences, emphasizing the importance of context in interpreting income trends. These insights are valuable for understanding global economic inequities in the data science field and can inform policies aimed at reducing disparities and promoting equitable professional development.
Q2


### Methods
Respondents' programming language usage and industry roles were derived from the survey's structured questions. Experience levels were classified into beginner, intermediate, and expert categories using the years of coding experience provided by the respondents. Proportional usage of programming languages across experience levels and industries was computed to identify trends. Visualizations were generated using stacked bar charts to compare the proportional adoption of programming languages across both experience levels and industry categories.


### Analysis
```{r}
selected_data <- data %>%
  select(Q6, Q7_Part_1:Q7_Part_12) %>%
  rename(
    Experience = Q6,
    Python = Q7_Part_1,
    R = Q7_Part_2,
    SQL = Q7_Part_3,
    C = Q7_Part_4,
    Cpp = Q7_Part_5,
    Java = Q7_Part_6,
    JavaScript = Q7_Part_7,
    Julia = Q7_Part_8,
    Swift = Q7_Part_9,
    Bash = Q7_Part_10,
    MATLAB = Q7_Part_11,
    None = Q7_Part_12
  )

selected_data <- selected_data %>%
  filter(Experience != "I have never written code")
```



```{r}


# Pivot table
pivot_data <- selected_data %>%
  pivot_longer(
    cols = Python:None, # Columns for programming languages
    names_to = "Language", # Create a new column for language names
    values_to = "Value"    # Create a new column for values (e.g., Yes/No or other indicators)
  ) %>%
  mutate(Value = ifelse(Value!='', 1, 0)) %>% # Convert "Yes" to 1 and others to 0
  group_by(Experience, Language) %>%
  summarize(Count = sum(Value, na.rm = TRUE), .groups = 'drop') %>%
  pivot_wider(
    names_from = Language, # Spread languages into columns
    values_from = Count,   # Fill with counts
    values_fill = 0        # Fill missing values with 0
  )

# Print the final pivot table
pivot_data <- pivot_data[-1,]
pivot_data

```

```{r}
create_pie_chart_with_legend <- function(column_name) {
  values <- pivot_data[[column_name]]
  labels <- pivot_data$Experience
  total <- sum(values)
  
  # Calculate percentages
  percentages <- round(values / total * 100, 1)
  legend_labels <- paste(labels, "(", percentages, "%)", sep = "")
  
  # Generate pie chart
  pie(values, 
      labels = NA, # Remove labels from the pie chart
      main = paste("Pie Chart for", column_name),
      col = rainbow(length(values)))
  
  # Add a legend
  legend("topright", legend = legend_labels, fill = rainbow(length(values)), cex = 0.8)
}

# Generate pie charts for each column
for (column in colnames(pivot_data)[-1]) {
  create_pie_chart_with_legend(column)
}

```




```{r}
pivot_data_long <- pivot_data %>%
  pivot_longer(cols = -Experience, names_to = "Language", values_to = "Usage")
pivot_data_long <- pivot_data_long %>%
  group_by(Language) %>%
  mutate(Proportion = Usage / sum(Usage))
pivot_data_long <- pivot_data_long %>%
  group_by(Language) %>%
  mutate(Proportion = Usage / sum(Usage))

# Create proportional stacked bar chart
ggplot(pivot_data_long, aes(x = Language, y = Proportion, fill = Experience)) +
  geom_bar(stat = "identity", position = "fill") +
  labs(title = "Proportional Programming Language Usage by Experience Level", 
       x = "Programming Language", 
       y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}
selected_data <- data %>%
  select(Q5, Q7_Part_1:Q7_Part_12) %>%
  rename(
    Industry = Q5,
    Python = Q7_Part_1,
    R = Q7_Part_2,
    SQL = Q7_Part_3,
    C = Q7_Part_4,
    Cpp = Q7_Part_5,
    Java = Q7_Part_6,
    JavaScript = Q7_Part_7,
    Julia = Q7_Part_8,
    Swift = Q7_Part_9,
    Bash = Q7_Part_10,
    MATLAB = Q7_Part_11,
    None = Q7_Part_12
  )
selected_data
```

```{r}
pivot_data <- selected_data %>%
  pivot_longer(
    cols = Python:None, # Columns for programming languages
    names_to = "Language", # Create a new column for language names
    values_to = "Value"    # Create a new column for values (e.g., Yes/No or other indicators)
  ) %>%
  mutate(Value = ifelse(Value!='', 1, 0)) %>% # Convert "Yes" to 1 and others to 0
  group_by(Industry, Language) %>%
  summarize(Count = sum(Value, na.rm = TRUE), .groups = 'drop') %>%
  pivot_wider(
    names_from = Language, # Spread languages into columns
    values_from = Count,   # Fill with counts
    values_fill = 0        # Fill missing values with 0
  )

# Print the final pivot table
pivot_data <- pivot_data[-1,]
pivot_data
```



```{r}
pivot_data_long <- pivot_data %>%
  pivot_longer(cols = -Industry, names_to = "Language", values_to = "Usage")
pivot_data_long <- pivot_data_long %>%
  group_by(Language) %>%
  mutate(Proportion = Usage / sum(Usage))
pivot_data_long <- pivot_data_long %>%
  group_by(Language) %>%
  mutate(Proportion = Usage / sum(Usage))

# Create proportional stacked bar chart
ggplot(pivot_data_long, aes(x = Language, y = Proportion, fill = Industry)) +
  geom_bar(stat = "identity", position = "fill") +
  labs(title = "Proportional Programming Language Usage by Experience Level", 
       x = "Programming Language", 
       y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}


# Define a larger, more varied color palette
color_palette <- brewer.pal(n = min(length(unique(pivot_data_long$Industry)), 12), "Set3")

# If more than 12 industries, expand the palette
if (length(unique(pivot_data_long$Industry)) > 12) {
  color_palette <- colorRampPalette(brewer.pal(12, "Set3"))(length(unique(pivot_data_long$Industry)))
}

# Plot with custom colors
ggplot(pivot_data_long, aes(x = Language, y = Proportion, fill = Industry)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = color_palette) +  # Apply the custom palette
  labs(title = "Proportional Programming Language Usage by Industry", 
       x = "Programming Language", 
       y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The analysis revealed distinct trends in programming language preferences among practitioners of different experience levels. Beginners showed a strong reliance on foundational languages like Python and SQL, while experts exhibited broader adoption of languages like R, Java, and Bash for advanced tasks. Across industries, Python emerged as a dominant language for data-related fields like data science and machine learning, whereas SQL had widespread adoption in roles such as data engineering and database management. More niche languages, such as Julia and MATLAB, saw usage concentrated in specialized roles like research scientists.


### Conclusion
The study revealed distinct programming language preferences that evolve with practitioners' experience levels and vary by industry roles. Python's widespread adoption across all levels underscores its utility as a foundational language in the data science ecosystem. Meanwhile, experts and industry specialists integrate niche languages to address specific challenges. These findings provide actionable insights for curriculum design in education, workforce training, and recruitment strategies, emphasizing the importance of Python for beginners and advanced language fluency for experienced professionals in specialized roles.



